---
title: "Getting Started with TaxonBodyMassML"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with TaxonBodyMassML}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE
)
```

## Installation

Install the package from the latest GitHub release with `pak` (the `latest`
release always points at the newest version; specific versions are listed on
the [releases page](https://github.com/TaxonBodyMassML/TaxonBodyMassML/releases)):

```{r install}
# install.packages("pak")
pak::pkg_install("url::https://github.com/TaxonBodyMassML/TaxonBodyMassML/releases/latest/download/TaxonBodyMassML_latest.tar.gz")
```

Model artifacts (~0.4 GB) are downloaded automatically the first time you call
`predict_mass()` or `lookup_taxonomy()`. You can also trigger the download
manually at any time:

```{r download}
TaxonBodyMassML::download_model()
```

## Quick Start

### Single species

```{r single}
library(TaxonBodyMassML)

TaxonBodyMassML::predict_mass("Homo sapiens")
#>          taxon mass_g
#> 1 Homo sapiens  66300
```

*Homo sapiens* has a recorded mass in the database, so the value returned is
that measurement rather than a model prediction (see Source Attribution below).

### Multiple species

```{r multiple}
TaxonBodyMassML::predict_mass(c("Canis lupus", "Panthera leo", "Mus musculus"))
#>          taxon    mass_g
#> 1  Canis lupus  10090.06
#> 2 Panthera leo 160500.00
#> 3  Mus musculus     18.02
```

### With confidence interval

```{r ci}
# 90% conformal prediction interval (default when confidence_interval = TRUE)
TaxonBodyMassML::predict_mass("Canis lupus", confidence_interval = TRUE)
#>         taxon   mass_g lower_bound upper_bound confidence
#> 1 Canis lupus 10090.06    1310.773    77671.24        0.9

# Custom coverage level
TaxonBodyMassML::predict_mass("Canis lupus", confidence_interval = 0.80)
#>         taxon   mass_g lower_bound upper_bound confidence
#> 1 Canis lupus 10090.06    2964.959    34337.52        0.8
```

### Include resolved taxonomy in output

```{r tax-output}
TaxonBodyMassML::predict_mass("Panthera leo", include_taxonomy = TRUE)
```

## Prediction Methods

Two models are shipped; both use only the taxonomy from kingdom to genus as
features (species is used for the training-data lookup and for identifying
genus-level queries, not as a model input).

| `method` | Description |
|---|---|
| `"EntityEmbeddings"` (default) | Two-stage model: learned per-rank taxonomy embeddings, then XGBoost on the embedding vectors. |
| `"XGBoost"` | Single XGBoost model with native categorical splits on the taxonomy ranks. |

```{r method}
TaxonBodyMassML::predict_mass("Nucella lima", method = "XGBoost")
#>          taxon   mass_g
#> 1 Nucella lima 1.583428
```

Both methods return the empirical mass for species present in the training
data unless `lookup = FALSE`.

Two kinds of input come back as `NA` with a warning: names that cannot be
resolved to a taxonomy at all, and names whose kingdom-to-genus classification
shares no value with the training data (for example plants and fungi, which
the training data exclude). Predicting the latter would mean extrapolating
from entirely unknown features, so no prediction is made and, with
`include_source = TRUE`, the `source` column reads `"tbmML_UNK"`.

## Source Attribution

Use `include_source = TRUE` to append a `source` column that records the
provenance of each returned mass value. For species that appear directly in the
training data, the column contains the original source identifier. For
model-inferred values, the column is `"tbmML_"` followed by the finest
taxonomic rank present in the training data.

```{r source}
TaxonBodyMassML::predict_mass(
  c("Nucella ostrina",  # in training data — returns empirical mass
    "Nucella lima"),    # not in training data — model infers from genus
  include_source = TRUE
)
#>             taxon   mass_g      source
#> 1 Nucella ostrina 0.700000  Novak_2017
#> 2    Nucella lima 1.341924 tbmML_genus
```

Note that for species in the training data, `mass_g` is the empirical measured
value (not a model prediction), and confidence interval columns (if requested)
will be `NA`. See `create_bib()` (under Citation, below) to export the cited
sources as a BibTeX file.

## Fuzzy Name Matching

`predict_mass()` can correct species names via the GBIF species-match API
before performing the taxonomy lookup, tolerating minor misspellings and name
variants. Set `fuzzy_match_name = TRUE` to enable this. A `matched_name`
column records the original query name when a correction is made; `taxon`
always holds the GBIF-canonical name used for prediction.

```{r fuzzy}
# Enable fuzzy correction — misspelled name is corrected automatically
TaxonBodyMassML::predict_mass("Canis lupis", fuzzy_match_name = TRUE)
#>         taxon   mass_g matched_name
#> 1 Canis lupus 10090.06  Canis lupis

# Default: exact name matching (no correction step)
TaxonBodyMassML::predict_mass("Canis lupus")
#>         taxon   mass_g
#> 1 Canis lupus 10090.06
```

`fuzzy_predict_mass()` is deprecated as of v0.2.1. Replace any calls to it
with `predict_mass(..., fuzzy_match_name = TRUE)`.

## Taxonomy Lookup

You can resolve taxonomy independently using `lookup_taxonomy()`. This returns
a `data.frame` that can be inspected or passed directly to `predict_mass()` to skip
the lookup step on subsequent calls.

```{r lookup}
tax <- TaxonBodyMassML::lookup_taxonomy(c("Homo sapiens", "Canis lupus"))
tax

# Pass pre-resolved taxonomy to predict_mass()
TaxonBodyMassML::predict_mass(tax)
```

Taxonomy resolution uses the [GBIF fuzzy-match API](https://www.gbif.org/developer/species)
with an [NCBI Entrez](https://www.ncbi.nlm.nih.gov/taxonomy) fallback.
Names that cannot be resolved in either service emit a `warning()` and receive
`NA` for all taxonomy columns and `NA` for `mass_g`.

## Confidence Intervals

`TaxonBodyMassML` uses conformal prediction to produce distribution-free
coverage guarantees calibrated on held-out residuals.

| `confidence_interval` value | Behaviour |
|---|---|
| `FALSE` (default) | No interval returned |
| `TRUE` | 90% interval |
| numeric in (0, 1) | Interval at that coverage level |

```{r ci-demo}
TaxonBodyMassML::predict_mass(
  c("Canis lupus", "Nucella lima"),  # neither has a recorded mass, so both are model predictions
  confidence_interval = 0.95
)
#>          taxon       mass_g lower_bound  upper_bound confidence
#> 1  Canis lupus 10090.062404  531.096793 191696.43012       0.95
#> 2 Nucella lima     1.341924    0.070633     25.49459       0.95
```

By default the interval is calibrated for the finest taxonomic rank that the
query resolved to (`interval_method = "stratified"`), so a query known only to
family or order level receives a correspondingly wider interval;
`interval_method = "pooled"` uses a single calibration set instead.

## Disk Cache

By default, resolved taxonomy is stored only in the in-memory session cache.
Enable persistent on-disk caching to avoid repeated network lookups across R
sessions:

```{r disk-cache}
TaxonBodyMassML::tbm_options(disk_cache = TRUE)

# Subsequent calls will read from / write to the cache directory
TaxonBodyMassML::lookup_taxonomy("Homo sapiens")

# Clear the disk cache when no longer needed
TaxonBodyMassML::tbm_clear_cache(disk = TRUE, session = FALSE)
```

The taxonomy cache is stored separately from the model artifacts in
`tools::R_user_dir("TaxonBodyMassML", "cache")/taxonomy/`.

## Package Options

```{r options}
# Get current options
TaxonBodyMassML::tbm_options()
#> $disk_cache
#> [1] FALSE
#>
#> $progress
#> [1] TRUE

# Set multiple options at once
TaxonBodyMassML::tbm_options(disk_cache = TRUE, progress = FALSE)
```

Available options:

| Option | Type | Default | Description |
|---|---|---|---|
| `disk_cache` | logical | `FALSE` | Cache taxonomy on disk across sessions |
| `progress` | logical | `TRUE` | Show progress bar for >10 species (requires `cli`) |

## API Compliance

### TAXONBODYMASSML_EMAIL

Set this environment variable to identify your application in the User-Agent
header sent to GBIF and NCBI:

```{r email}
Sys.setenv(TAXONBODYMASSML_EMAIL = "your.email@example.com")
```

Or add it to your `~/.Renviron`:

```
TAXONBODYMASSML_EMAIL=your.email@example.com
```

### NCBI_API_KEY

NCBI rate-limits unauthenticated requests to 3 per second. Register for a
free API key at <https://account.ncbi.nlm.nih.gov/> to increase the
limit to 10 requests per second:

```{r ncbi-key}
Sys.setenv(NCBI_API_KEY = "your_ncbi_api_key")
```

## Data Sources

Training data on each taxon's mass (in grams) was obtained from unpublished
primary data, published allometric relationships, the primary literature, and
published databases. Most mass data represent mean adult wet mass. For fishes
for which Fishbase did not report mass but did report body length, we used the
allometric coefficients of Froese et al. (2014) to convert length to mass. For
lizards, we used the maximum snout-ventral lengths and allometric coefficients
of Meiri (2018). Published mass data came from Meiri (2018), Oliveira et al.
(2017), Brown et al. (2018), Smith et al. (2003), Anderson et al. (2017),
Gillooly et al. (2016), Jennings et al. (2002), Lislevand et al. (2007),
Killen et al. (2016), Feldman et al. (2016), Tucker et al. (2014a, b), Hirt et
al. (2017), Eklöf et al. (2017), Cai et al. (2025), Animal Diversity Web (via
Qaardvark), AnAge (Tacutu et al., 2013), Fishbase (Froese and Pauly, 2025),
Sealifebase (Palomares and Pauly, 2025), and DataRetriever (including
mammal-life-hist, bird-size, home-ranges, amniote-life-hist, socean-diet-data,
vertnet-amphibian, vertnet-reptiles; McGlinn et al. 2017).

### References

Anderson, D.M. and Gillooly, J.F. (2017). Physiological constraints on
long-term population cycles: a broad-scale view. *Evolutionary Ecology
Research*, 18(6), 693–707.

Brown, J.H., Hall, C.A.S. and Sibly, R.M. (2018). Equal fitness paradigm
explained by a trade-off between generation time and energy production rate.
*Nature Ecology & Evolution*, 2(2), 262–268.
<https://doi.org/10.1038/s41559-017-0430-1>

Cai, T., Wen, Z., Jiang, Z. and Zhen, Y. (2025). Distinct latitudinal patterns
of molecular rates across vertebrates. *Proceedings of the National Academy of
Sciences*, 122(19), e2423386122.
<https://doi.org/10.1073/pnas.2423386122>

Eklöf, J., Austin, Å., Bergström, U., Donadi, S., Eriksson, B.K., Hansen, J.
and Sundblad, G. (2017). Size matters: relationships between body size and body
mass of common coastal, aquatic invertebrates in the Baltic Sea. *PeerJ*, 5,
e2906.

Feldman, A., Sabath, N., Pyron, R.A., Mayrose, I. and Meiri, S. (2016). Body
sizes and diversification rates of lizards, snakes, amphisbaenians and the
tuatara. *Global Ecology and Biogeography*, 25(2), 187–197.

Froese, R. and Pauly, D. (Eds.) (2025). FishBase. World Wide Web electronic
publication. <https://www.fishbase.org>

Froese, R., Thorson, J.T. and Reyes Jr, R.B. (2014). A Bayesian approach for
estimating length-weight relationships in fishes. *Journal of Applied
Ichthyology*, 30(1), 78–85.
<https://doi.org/10.1111/jai.12299>

Gillooly, J.F., Gomez, J.P., Mavrodiev, E.V., Rong, Y. and McLamore, E.S.
(2016). Body mass scaling of passive oxygen diffusion in endotherms and
ectotherms. *Proceedings of the National Academy of Sciences*, 113(19),
5340–5345.

Hirt, M.R., Jetz, W., Rall, B.C. and Brose, U. (2017). A general scaling law
reveals why the largest animals are not the fastest. *Nature Ecology &
Evolution*, 1(8), 1116–1122.

Jennings, S., Pinnegar, J.K., Polunin, N.V.C. and Warr, K.J. (2002). Linking
size-based and trophic analyses of benthic community structure. *Marine Ecology
Progress Series*, 226, 77–85.

Killen, S.S., Glazier, D.S., Rezende, E.L., Clark, T.D., Atkinson, D.,
Willener, A.S.T. and Halsey, L.G. (2016). Ecological influences and
morphological correlates of resting and maximal metabolic rates across teleost
fish species. *The American Naturalist*, 187(5), 592–606.

Lislevand, T., Figuerola, J. and Székely, T. (2007). Avian body sizes in
relation to fecundity, mating system, display behavior, and resource sharing:
Ecological Archives E088-096. *Ecology*, 88(6), 1605.

McGlinn, D.J., Senyondo, H., Sharma, S., Ye, H., Taylor, S.D., Ooms, J.,
Goel, N., Velazco, S.J.E., White, E.P., Baldridge, E. and Michonneau, F.
(2017). retriever: R Interface to the Data Retriever. *Journal of Open Source
Software*, 2(17), 451.
<https://doi.org/10.21105/joss.00451>

Meiri, S. (2018). Traits of lizards of the world: Variation around a
successful evolutionary design. *Global Ecology and Biogeography*, 27(10),
1168–1172.

Oliveira, B.F., São-Pedro, V.A., Santos-Barrera, G., Penone, C. and Costa,
G.C. (2017). AmphiBIO, a global database for amphibian ecological traits.
*Scientific Data*, 4(1), 170123.
<https://doi.org/10.1038/sdata.2017.123>

Palomares, M.L.D. and Pauly, D. (Eds.) (2025). SeaLifeBase. World Wide Web
electronic publication. <https://www.sealifebase.ca>

Smith, F.A., Lyons, S.K., Ernest, S.K.M., Jones, K.E., Kaufman, D.M., Dayan,
T., Marquet, P.A., Brown, J.H. and Haskell, J.P. (2003). Body mass of late
Quaternary mammals. *Ecology*, 84(12), 3403–3403.
<https://doi.org/10.1890/02-9003>

Tacutu, R., Craig, T., Budovsky, A., Wuttke, D., Lehmann, G., Taranukha, D.,
Costa, J., Fraifeld, V.E. and de Magalhães, J.P. (2013). Human Ageing Genomic
Resources: Integrated databases and tools for the biology and genetics of
ageing. *Nucleic Acids Research*, 41(D1), D1027–D1033.

Tucker, M.A. and Rogers, T.L. (2014a). Examining predator–prey body size,
trophic level and body mass across marine and terrestrial mammals. *Proceedings
of the Royal Society B: Biological Sciences*, 281(1797), 20142103.

Tucker, M.A., Ord, T.J. and Rogers, T.L. (2014b). Evolutionary predictors of
mammalian home range size: body mass, diet and the environment. *Global Ecology
and Biogeography*, 23(10), 1105–1114.

## Citation

If you use TaxonBodyMassML in your work, please cite it:

```{r citation}
citation("TaxonBodyMassML")
```

The bibliography for the underlying body-mass data sources is available at:

```{r bib-path}
TaxonBodyMassML::get_citations()
```

This path can be loaded with `bibtex::read.bib()` (requires the `bibtex`
package):

```{r bibtex}
refs <- bibtex::read.bib(TaxonBodyMassML::get_citations())
```

To cite only the sources behind a particular set of predictions, pass the
result of `predict_mass(..., include_source = TRUE)` to `create_bib()`. It
writes a `.bib` file containing the entries for every training-data source
named in the `source` column; model-inferred rows (`tbmML_*`) are skipped.

```{r create-bib}
res <- TaxonBodyMassML::predict_mass(
  c("Nucella ostrina", "Anolis carolinensis", "Nucella lima"),
  include_source = TRUE
)
TaxonBodyMassML::create_bib(res, file = "my_sources.bib")
```
