---
title: "Getting Started with TaxonBodyMassML"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with TaxonBodyMassML}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE
)
```

## Installation

Install the package from the latest release. Copy the `.tar.gz` URL from the
[GitHub releases page](https://github.com/TaxonBodyMassML/TaxonBodyMassML/releases)
and pass it to `pak`:

```{r install}
# install.packages("pak")
pak::pkg_install("url::<release-url>")
```

Model artifacts (~2 GB) are downloaded automatically the first time you call
`predict_mass()` or `lookup_taxonomy()`. You can also trigger the download
manually at any time:

```{r download}
TaxonBodyMassML::download_model()
```

## Quick Start

### Single species

```{r single}
library(TaxonBodyMassML)

TaxonBodyMassML::predict_mass("Homo sapiens")
#>          taxon   mass_g
#> 1 Homo sapiens 73512.41
```

### Multiple species

```{r multiple}
TaxonBodyMassML::predict_mass(c("Canis lupus", "Panthera leo", "Mus musculus"))
#>          taxon     mass_g
#> 1  Canis lupus  31452.21
#> 2 Panthera leo 163241.87
#> 3  Mus musculus     20.14
```

### With confidence interval

```{r ci}
# 90% conformal prediction interval (default when confidence_interval = TRUE)
TaxonBodyMassML::predict_mass("Canis lupus", confidence_interval = TRUE)

# Custom coverage level
TaxonBodyMassML::predict_mass("Canis lupus", confidence_interval = 0.80)
```

### Include resolved taxonomy in output

```{r tax-output}
TaxonBodyMassML::predict_mass("Panthera leo", include_taxonomy = TRUE)
```

## Fuzzy Name Matching

`predict_mass()` can correct species names via the GBIF species-match API
before performing the taxonomy lookup, tolerating minor misspellings and name
variants. Set `fuzzy_match_name = TRUE` to enable this. A `matched_name`
column records the original query name when a correction is made; `taxon`
always holds the GBIF-canonical name used for prediction.

```{r fuzzy}
# Enable fuzzy correction — misspelled name is corrected automatically
TaxonBodyMassML::predict_mass("Canis lupuss", fuzzy_match_name = TRUE)
#>         taxon   mass_g matched_name
#> 1 Canis lupus 31452.21 Canis lupuss

# Default: exact name matching (no correction step)
TaxonBodyMassML::predict_mass("Canis lupus")
#>         taxon   mass_g
#> 1 Canis lupus 31452.21
```

`fuzzy_predict_mass()` is deprecated as of v0.2.1. Replace any calls to it
with `predict_mass(..., fuzzy_match_name = TRUE)`.

## Taxonomy Lookup

You can resolve taxonomy independently using `lookup_taxonomy()`. This returns
a `data.frame` that can be inspected or passed directly to `predict_mass()` to skip
the lookup step on subsequent calls.

```{r lookup}
tax <- TaxonBodyMassML::lookup_taxonomy(c("Homo sapiens", "Canis lupus"))
tax

# Pass pre-resolved taxonomy to predict_mass()
TaxonBodyMassML::predict_mass(tax)
```

Taxonomy resolution uses the [GBIF fuzzy-match API](https://www.gbif.org/developer/species)
with an [NCBI Entrez](https://www.ncbi.nlm.nih.gov/taxonomy) fallback.
Names that cannot be resolved in either service emit a `warning()` and receive
`NA` for all taxonomy columns and `NA` for `mass_g`.

## Confidence Intervals

`TaxonBodyMassML` uses conformal prediction to produce distribution-free
coverage guarantees calibrated on held-out residuals.

| `confidence_interval` value | Behaviour |
|---|---|
| `FALSE` (default) | No interval returned |
| `TRUE` | 90% interval |
| numeric in (0, 1) | Interval at that coverage level |

```{r ci-demo}
TaxonBodyMassML::predict_mass(
  c("Homo sapiens", "Mus musculus"),
  confidence_interval = 0.95
)
#>          taxon    mass_g lower_bound upper_bound confidence
#> 1 Homo sapiens 73512.41    4823.61   1120423.2       0.95
#> 2  Mus musculus    20.14       1.32       307.1       0.95
```

## Disk Cache

By default, resolved taxonomy is stored only in the in-memory session cache.
Enable persistent on-disk caching to avoid repeated network lookups across R
sessions:

```{r disk-cache}
TaxonBodyMassML::tbm_options(disk_cache = TRUE)

# Subsequent calls will read from / write to the cache directory
TaxonBodyMassML::lookup_taxonomy("Homo sapiens")

# Clear the disk cache when no longer needed
TaxonBodyMassML::tbm_clear_cache(disk = TRUE, session = FALSE)
```

The taxonomy cache is stored separately from the model artifacts in
`tools::R_user_dir("TaxonBodyMassML", "cache")/taxonomy/`.

## Package Options

```{r options}
# Get current options
TaxonBodyMassML::tbm_options()
#> $disk_cache
#> [1] FALSE
#>
#> $progress
#> [1] TRUE

# Set multiple options at once
TaxonBodyMassML::tbm_options(disk_cache = TRUE, progress = FALSE)
```

Available options:

| Option | Type | Default | Description |
|---|---|---|---|
| `disk_cache` | logical | `FALSE` | Cache taxonomy on disk across sessions |
| `progress` | logical | `TRUE` | Show progress bar for >10 species (requires `cli`) |

## API Compliance

### TAXONBODYMASSML_EMAIL

Set this environment variable to identify your application in the User-Agent
header sent to GBIF and NCBI:

```{r email}
Sys.setenv(TAXONBODYMASSML_EMAIL = "your.email@example.com")
```

Or add it to your `~/.Renviron`:

```
TAXONBODYMASSML_EMAIL=your.email@example.com
```

### NCBI_API_KEY

NCBI rate-limits unauthenticated requests to 3 per second. Register for a
free API key at <https://account.ncbi.nlm.nih.gov/> to increase the
limit to 10 requests per second:

```{r ncbi-key}
Sys.setenv(NCBI_API_KEY = "your_ncbi_api_key")
```

## Data Sources

Training data on each taxon's mass (in grams) was obtained from unpublished
primary data, published allometric relationships, the primary literature, and
published databases. Most mass data represent mean adult wet mass. For fishes
for which Fishbase did not report mass but did report body length, we used the
allometric coefficients of Froese et al. (2014) to convert length to mass. For
lizards, we used the maximum snout-ventral lengths and allometric coefficients
of Meiri (2018). Published mass data came from Meiri (2018), Oliveira et al.
(2017), Brown et al. (2018), Smith et al. (2003), Anderson et al. (2017),
Gillooly et al. (2016), Jennings et al. (2002), Lislevand et al. (2007),
Killen et al. (2016), Feldman et al. (2016), Tucker et al. (2014a, b), Hirt et
al. (2017), Eklöf et al. (2017), Cai et al. (2025), Animal Diversity Web (via
Qaardvark), AnAge (Tacutu et al., 2013), Fishbase (Froese and Pauly, 2025),
Sealifebase (Palomares and Pauly, 2025), and DataRetriever (including
mammal-life-hist, bird-size, home-ranges, amniote-life-hist, socean-diet-data,
vertnet-amphibian, vertnet-reptiles; McGlinn et al. 2017).

### References

Anderson, D.M. and Gillooly, J.F. (2017). Physiological constraints on
long-term population cycles: a broad-scale view. *Evolutionary Ecology
Research*, 18(6), 693–707.

Brown, J.H., Hall, C.A.S. and Sibly, R.M. (2018). Equal fitness paradigm
explained by a trade-off between generation time and energy production rate.
*Nature Ecology & Evolution*, 2(2), 262–268.
<https://doi.org/10.1038/s41559-017-0430-1>

Cai, T., Wen, Z., Jiang, Z. and Zhen, Y. (2025). Distinct latitudinal patterns
of molecular rates across vertebrates. *Proceedings of the National Academy of
Sciences*, 122(19), e2423386122.
<https://doi.org/10.1073/pnas.2423386122>

Eklöf, J., Austin, Å., Bergström, U., Donadi, S., Eriksson, B.K., Hansen, J.
and Sundblad, G. (2017). Size matters: relationships between body size and body
mass of common coastal, aquatic invertebrates in the Baltic Sea. *PeerJ*, 5,
e2906.

Feldman, A., Sabath, N., Pyron, R.A., Mayrose, I. and Meiri, S. (2016). Body
sizes and diversification rates of lizards, snakes, amphisbaenians and the
tuatara. *Global Ecology and Biogeography*, 25(2), 187–197.

Froese, R. and Pauly, D. (Eds.) (2025). FishBase. World Wide Web electronic
publication. <https://www.fishbase.org>

Froese, R., Thorson, J.T. and Reyes Jr, R.B. (2014). A Bayesian approach for
estimating length-weight relationships in fishes. *Journal of Applied
Ichthyology*, 30(1), 78–85.
<https://doi.org/10.1111/jai.12299>

Gillooly, J.F., Gomez, J.P., Mavrodiev, E.V., Rong, Y. and McLamore, E.S.
(2016). Body mass scaling of passive oxygen diffusion in endotherms and
ectotherms. *Proceedings of the National Academy of Sciences*, 113(19),
5340–5345.

Hirt, M.R., Jetz, W., Rall, B.C. and Brose, U. (2017). A general scaling law
reveals why the largest animals are not the fastest. *Nature Ecology &
Evolution*, 1(8), 1116–1122.

Jennings, S., Pinnegar, J.K., Polunin, N.V.C. and Warr, K.J. (2002). Linking
size-based and trophic analyses of benthic community structure. *Marine Ecology
Progress Series*, 226, 77–85.

Killen, S.S., Glazier, D.S., Rezende, E.L., Clark, T.D., Atkinson, D.,
Willener, A.S.T. and Halsey, L.G. (2016). Ecological influences and
morphological correlates of resting and maximal metabolic rates across teleost
fish species. *The American Naturalist*, 187(5), 592–606.

Lislevand, T., Figuerola, J. and Székely, T. (2007). Avian body sizes in
relation to fecundity, mating system, display behavior, and resource sharing:
Ecological Archives E088-096. *Ecology*, 88(6), 1605.

McGlinn, D.J., Senyondo, H., Sharma, S., Ye, H., Taylor, S.D., Ooms, J.,
Goel, N., Velazco, S.J.E., White, E.P., Baldridge, E. and Michonneau, F.
(2017). retriever: R Interface to the Data Retriever. *Journal of Open Source
Software*, 2(17), 451.
<https://doi.org/10.21105/joss.00451>

Meiri, S. (2018). Traits of lizards of the world: Variation around a
successful evolutionary design. *Global Ecology and Biogeography*, 27(10),
1168–1172.

Oliveira, B.F., São-Pedro, V.A., Santos-Barrera, G., Penone, C. and Costa,
G.C. (2017). AmphiBIO, a global database for amphibian ecological traits.
*Scientific Data*, 4(1), 170123.
<https://doi.org/10.1038/sdata.2017.123>

Palomares, M.L.D. and Pauly, D. (Eds.) (2025). SeaLifeBase. World Wide Web
electronic publication. <https://www.sealifebase.ca>

Smith, F.A., Lyons, S.K., Ernest, S.K.M., Jones, K.E., Kaufman, D.M., Dayan,
T., Marquet, P.A., Brown, J.H. and Haskell, J.P. (2003). Body mass of late
Quaternary mammals. *Ecology*, 84(12), 3403–3403.
<https://doi.org/10.1890/02-9003>

Tacutu, R., Craig, T., Budovsky, A., Wuttke, D., Lehmann, G., Taranukha, D.,
Costa, J., Fraifeld, V.E. and de Magalhães, J.P. (2013). Human Ageing Genomic
Resources: Integrated databases and tools for the biology and genetics of
ageing. *Nucleic Acids Research*, 41(D1), D1027–D1033.

Tucker, M.A. and Rogers, T.L. (2014a). Examining predator–prey body size,
trophic level and body mass across marine and terrestrial mammals. *Proceedings
of the Royal Society B: Biological Sciences*, 281(1797), 20142103.

Tucker, M.A., Ord, T.J. and Rogers, T.L. (2014b). Evolutionary predictors of
mammalian home range size: body mass, diet and the environment. *Global Ecology
and Biogeography*, 23(10), 1105–1114.

## Citation

If you use TaxonBodyMassML in your work, please cite it:

```{r citation}
citation("TaxonBodyMassML")
```

The bibliography for the underlying body-mass data sources is available at:

```{r bib-path}
TaxonBodyMassML::get_citations()
```

This path can be loaded with `bibtex::read.bib()` (requires the `bibtex`
package):

```{r bibtex}
refs <- bibtex::read.bib(TaxonBodyMassML::get_citations())
```
